First Order Alternation

### 2018-11-06 quantifiers/body.tex
ok

\documentclass{llncs}
% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}

\usepackage[usenames,dvipsnames]{color}

\usepackage{latexsym}
\usepackage{amsxtra} 
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{paralist}
% \usepackage{stmaryrd}
\usepackage{txfonts}
\usepackage{framed}
\usepackage{makecell}
\usepackage{url}
\usepackage{tikz}
% \usepackage{hyperref}
\usetikzlibrary{automata,positioning, calc}
\usepackage[inline,shortlabels]{enumitem}

\usepackage{proof}

\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}

\usepackage[draft]{commenting}

\pagestyle{plain}

\include{commands-lncs}

\declareauthor{ri}{Radu}{blue}
\declareauthor{xx}{Xiao}{red}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{First Order Alternation}

\author{Radu Iosif and Xiao Xu}
\institute{
  CNRS, Verimag, Universit\'e de Grenoble Alpes \\
  Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr
}

\maketitle

\begin{abstract}
We introduce first order alternating automata, a generalization of
boolean alternating automata, in which transition rules are described
by multisorted first order formulae, with states and internal
variables given by uninterpreted predicate terms. The model is closed
under union, intersection and complement, and its emptiness problem is
undecidable, even for the simplest data theory of equality. To cope
with this limitation, we develop an abstraction refinement
semi-algorithm based on lazy annotation of the symbolic execution
paths with interpolants, obtained by applying (i) quantifier
elimination with witness term generation and (ii) Lyndon interpolation
in the quantifier-free data theory with uninterpreted predicate
symbols. This provides a method for checking inclusion of timed and
finite-memory register automata, and emptiness of quantified predicate
automata, previously used in the verification of parameterized
concurrent programs, composed of replicated threads, with a
shared-memory communication model.
\end{abstract}

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
case we say that $\phi$ \emph{entails} $\psi$.

@@

\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
Observe that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

@@

We check emptiness of first order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(booleans) and no transition quantifiers. For simplicity, we do not
present the details of this algorithm (see Appendix
\ref{app:pseudocode}) and shall content ourselves of several
high-level definitions.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer the reader to \cite{IosifXu18} (or Appendix
\ref{app:pseudocode}).

@@

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{c}
q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
\arrow{(a,\nu_n)}{} \emptyset
\end{array}\]
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
formula using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

@@

\begin{theorem}\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence, 
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let: 
  \[\eta_i \equiv \left\{\begin{array}{ll}
  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i = \forall \\
  x_i & \text{if } Q_i = \exists
  \end{array}\right.\]
  where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
    y_{k_i}} = \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1
  x_1 \ldots Q_n x_n ~.~ \phi \models
  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) holds. 
\end{theorem}
\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
yields the unsatisfiable conjunction: 
\[\begin{array}{c}
z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
\end{array}\]
A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
\end{example}

@@

    \[\begin{array}{l}   
    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
    \end{array}\]

@@

    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq

@@

      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]

@@

\section{Applications}
\label{sec:applications}

@@

As before, this result enables using the
semi-algorithm from \S\ref{sec:emptiness} for checking emptiness of
predicate automata.  We point out that, although quantifier-free
predicate automata with predicates of arity one are decidable for
emptiness \cite{Farzan15}, currently there is no method for checking
emptiness of predicate automata with predicates of arity greater than
one, other than the explicit enumeration of cubes. Moreover, no method
for dealing with emptiness in the presence of transition quantifiers
is known to exist.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{abbrv} \bibliography{refs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as usual. 
\ifLongVersion
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere.  
\fi 
For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
for sentences. A sentence $\phi$ is \emph{satisfiable}
(\emph{unsatisfiable}) if $\sem{\phi}\neq\emptyset$
($\sem{\phi}=\emptyset$). An element of $\sem{\phi}$ is called a
\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
\models \phi$ for every interpretation $\I$ and every valuation
$\nu$. For two formulae $\phi$ and $\psi$ we write $\phi \models \psi$
for $\sem{\phi} \subseteq \sem{\psi}$, in which case we say that
$\phi$ \emph{entails} $\psi$.

@@

\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \vspace*{-.5\baselineskip}
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  
      \vspace*{-.5\baselineskip}
      We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
Observe that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

@@

We check emptiness of first order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(booleans) and no transition quantifiers. For simplicity, we do not
present the details of this algorithm and shall content ourselves of
several high-level definitions.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer the reader to \cite{IosifXu18}.

@@

\paragraph{Remark}
When considering an alternating model of computation, control-local
interpolants are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
sequence of input values. Consider, for instance, an automaton with
the following transition rules and final state $q_f$:
\[\begin{array}{rclcrclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) & \arrow{a(x)}{} & y+x > 0 \wedge q_f & \hspace*{2mm} & 
q_1(y) & \arrow{a(x)}{} & q_1(y+x) \\ 
q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f && 
q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{c}
q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
\arrow{(a,\nu_n)}{} \emptyset
\end{array}\]
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
formula using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

@@

\begin{theorem}\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
  holds.
\end{theorem}
\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
yields the unsatisfiable conjunction: 
\vspace*{-\baselineskip}
\[\begin{array}{c}
z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  ~\wedge~
\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) ~\wedge~ \\
\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
\end{array}\]
A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
\end{example}

@@

    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where \(\set{i_1,
    \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}\) and
    \(\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
    \set{x_k \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]\)

@@

    \ldots, I_n)$ for $\alpha$, such that \(\voc{I_k} \subseteq

@@

      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\).

@@

\ifLongVersion
\section{Applications}
\label{sec:applications}

@@

As before, this result enables using the
semi-algorithm from \S\ref{sec:emptiness} for checking emptiness of
predicate automata.  We point out that, although quantifier-free
predicate automata with predicates of arity one are decidable for
emptiness \cite{Farzan15}, currently there is no method for checking
emptiness of predicate automata with predicates of arity greater than
one, other than the explicit enumeration of cubes. Moreover, no method
for dealing with emptiness in the presence of transition quantifiers
is known to exist.
\fi
